---
title: "Modeling AutoClaims with tweedie Package"
author: Seth Temple
output: html_notebook
---


```{r}
library(tweedie)
library(statmod)
library(ggplot2)
library(insuranceData)
```

```{r}
data(AutoClaims)
```


Wrangle YEAR column.
```{r}
Year <- with(AutoClaims, {
  (AGE <= 99 & AGE >= 90)*9 + (AGE < 90 & AGE >= 80)*8 +
    (AGE < 80 & AGE >= 70)*7 + (AGE < 70 & AGE >= 60)*6 +
    (AGE < 60 & AGE >= 50)*5
})
AutoClaims$YEAR <- as.factor(Year)
rm(Year)
```

Build a model.

Model with no features.
```{r}
m0 <- glm(PAID~1, data=AutoClaims, family=tweedie(link.power=0, var.power=2))
summary(m0)
```

Add a feature.
```{r}
m1 <- glm(PAID~1+YEAR, data=AutoClaims, family=tweedie(link.power=0,var.power=2))
summary(m1)
anova(m1,m0)

print("AIC change")
print(AICtweedie(m1) - AICtweedie(m0))
```
We decreased deviance and AIC by a little.

Try to add more features.
```{r}
m2 <- glm(PAID~1+YEAR+CLASS, data=AutoClaims, family=tweedie(link.power=0,var.power=2))
summary(m2)
anova(m2,m1)

print("AIC change")
print(AICtweedie(m2) - AICtweedie(m1))
```

```{r}
m3 <- glm(PAID~1+YEAR+CLASS+STATE, data=AutoClaims,
          family=tweedie(link.power=0,var.power=2))
summary(m3)
anova(m3,m2)

print("AIC change")
print(AICtweedie(m3) - AICtweedie(m2))
```
We'll use features STATE, CLASS, and YEAR. The one feature we don't use is GENDER. We got an increase in AIC with GENDER included.

Use tweedie.profile to find estimate for p.
```{r}
start <- Sys.time()

# identity link
output <- tweedie.profile(PAID~1, data=AutoClaims, link.power=1,
                          method="interpolation", verbose=2,
                          do.plot=TRUE, do.smooth=FALSE,
                          p.vec=seq(2,3,.1))

end <- Sys.time()
print(end-start)
```

```{r}
start <- Sys.time()

# identity link
output <- tweedie.profile(PAID~1, data=AutoClaims, link.power=1,
                          method="interpolation", verbose=2,
                          do.plot=TRUE, do.smooth=FALSE,
                          p.vec=seq(2,3,.05))

end <- Sys.time()
print(end-start)
```

```{r}
start <- Sys.time()

# identity link
output <- tweedie.profile(PAID~1, data=AutoClaims, link.power=1,
                          method="interpolation", verbose=2,
                          do.plot=TRUE, do.smooth=FALSE,
                          p.vec=seq(2,3,.025))

end <- Sys.time()
print(end-start)
```
Our approach says that p=2.55 and phi=.02243302. Notably, phi is very small. This probably has to do with the fact that mu is larger in this dataset. Variance is a function of mu, phi, and p.

Compare the gamma model to the Tweedie model with p=2.55.

Write some functions to get feedback.
```{r}
give_feedback <- function(model){
  SSR <- sum((model$residuals)**2)
  fitted_values <- model$fitted.values
  print("Squared Sum of Residuals"); print(SSR)
  print("Deviance"); print(model$deviance)
  print("AIC"); print(AICtweedie(model))
}

compare_insurance_models <- function(model1, model2){
  
  SSR1 <- sum((model1$residuals)**2)
  SSR2 <- sum((model2$residuals)**2)
  diff_SSR <- SSR1 - SSR2
  
  fitted_values1 <- model1$fitted.values
  fitted_values2 <- model2$fitted.values
  diff_fitted_values <- fitted_values1 - fitted_values2
  diff_losses <- sum(abs(diff_fitted_values))
  diff_econ = sum(model1$fitted.values - model2$fitted.values)
  
  diff_deviance <- model1$deviance - model2$deviance
  diff_AIC <- AICtweedie(model1) - AICtweedie(model2)
  
  print("Model differences are relative to the first model")
  print("Squared Sum of Residuals"); print(diff_SSR)
  print("Deviance"); print(diff_deviance)
  print("AIC"); print(diff_AIC)
  
  print("Sum of absolute differences between losses"); print(diff_losses)
  print("Total difference between losses"); print(diff_econ)
  
  print("First 10 differences in fitted losses")
  print(diff_fitted_values[1:5])
  
}
```

The gamma model is m3.
```{r}
give_feedback(m3)
```


Make the Tweedie model with p=2.55.
```{r}
m4 <- glm(PAID~1+YEAR+CLASS+STATE, data=AutoClaims,
          family=tweedie(link.power=0,var.power=2.55))
give_feedback(m4)
summary(m4)
```

Make the inverse-Gaussian model (p=3).
```{r}
m5 <- glm(PAID~1+YEAR+CLASS+STATE, data=AutoClaims,
          family=tweedie(link.power=0,var.power=3))
give_feedback(m5)
summary(m5)
```

Compare p=2 and p=2.55 models.
```{r}
compare_insurance_models(m4,m3)
```

Compare p=2.55 and p=3 models.
```{r}
compare_insurance_models(m4,m5)
```

Compare gamma and inverse-Gaussian models.
```{r}
compare_insurance_models(m5,m3)
```

Take at a sneak peek at how fitted values change as p changes.
```{r}
m3$fitted.values[1:5]
m4$fitted.values[1:5]
m5$fitted.values[1:5]
```

Some observations.
- Tweedie model with p=2.55 estimates a dispersion parameter close to what tweedie.profile gives. The gamm model estimates phi close to 2.
- Tweedie model with p=2.55 has lower deviance and lower AIC.
- We could improve our predictions by ~$38,777.
- Differences in losses for individual observations is in the 10s.
- We only have ~6K observations.
- We get weird results for the deviance of the inverse-Gaussian model.

How much impact could changing the power parameter do to the modeling?
```{r}
38777/sum(AutoClaims$PAID)
```
Improving the model would barely change the total costs relative to how much loss is present. (I won't highlight this notebook in my thesis.)

Again, optimizing the power parameter provided little change in the grand scheme of things.
```{r}
sum(m3$fitted.values)/sum(AutoClaims$PAID)
sum(m4$fitted.values)/sum(AutoClaims$PAID)
sum(m5$fitted.values)/sum(AutoClaims$PAID)
```